---
title: "Introduction to sensiPhy"
author: "Gustavo Paterno; Caterine Penone; Gijsbert Werner"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: default
    highlight: pygments
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(fig.height = 6.5, fig.width = 10)
```

<br><br> 

### Sensitivity Analysis for Comparative Methods

The __sensiPhy__ package provides simple functions to perform sensitivity analysis
in phylogenetic comparative methods. It uses several simulation methods to 
estimate the impact of different types of uncertainty on PGLS models:  

(i) Sampling uncertainty (sample size; influential species and clades)  
(ii) Phylogenetic uncertainty    
(iii) Data uncertainty (intraspecific variation and measurement error)  

sensiPhy functions use a common syntax that combines the type of uncertainty and the type of model: 
*uncertainty.type_phylm* (for linear regressions) or *uncertainty.type_phyglm* (for logistic regressions).   

 **Function** | **Uncertainty** | **Model**  
----------- | ----------- | -----------
 samp_phylm | sample size | linear regression 
 samp_phyglm | sample size | logistic regression 
 tree_phylm | phylogenetic uncertain | linear regression 
 tree_phylm | phylogenetic uncertain | logistic regression 
 intra_phylm | Data uncertainty | linear regression 
 intra_phylm | Data uncertainty | logistic regression 

Mandatory arguments of sensiPhy functions include: `function(formula , data, phy, ...)`  
The following examples describes the basic usage of __sensiPhy__ functions. 

Loading the package and data 
```{r message=T}
set.seed(12345)
library(sensiPhy)

### Loading data:
data(alien)
data(primates) # se ?alien & ?primates for details about the data.
```

<br><br> 

## 1.Sampling uncertainty

### 1.1 Sensitivity analysis for sampling size:

The `samp_phylm` function performs analyses of sensitivity to species sampling by 
randomly removing species and detecting the effects on parameter estimates in 
a phylogenetic linear regression.  

* Additional arguments:   
__breaks:__ A vector containing the percentages of species to remove  
__times:__ The number of times species are randomly deleted for each break  

```{r samp_analysis, echo=T, cache=T, warning=FALSE}
# run analysis:
samp <- samp_phylm(log(gestaLen) ~ log(adultMass), phy = alien$phy[[1]], 
                   data = alien$data, times = 10, track = F)

# You can change the number of repetitions and break intervals:
samp2 <- samp_phylm(log(gestaLen) ~ log(adultMass), phy = alien$phy[[1]], track = F,
                    data = alien$data, times = 10, breaks = c(0.1, 0.2, 0.3, 0.4))
# You can change the phylogenetic model:
samp <- samp_phylm(log(gestaLen) ~ log(adultMass), phy = alien$phy[[1]], 
                   data = alien$data, model = "kappa", track = F)

# Check results:
knitr::kable(summary(samp))
# Visual diagnostics
sensi_plot(samp2)
# You can specify which graph and parameter ("slope" or "intercept") to print: 
sensi_plot(samp2, graphs = 1)
sensi_plot(samp2, param = "intercept")
```

<br><br>

### 1.2 Sensitivity analysis for influential species:

The function `influ_phylm` performs leave-one-out deletion analyis for 
phylogenetic linear regression, and detects influential species.

```{r influ_analysis, echo=T, cache=T, warning=FALSE}
# run analysis:
influ <- influ_phylm(log(gestaLen) ~ log(adultMass), phy = alien$phy[[1]], 
                     data = alien$data, track = F)
# To check summary results:
summary(influ)
# Most influential species
influ$influential.species
# Visual diagnostics
sensi_plot(influ)

# Check most influential species on the original regression plot:
sensi_plot(influ, graphs = 2)
```

<br><br>

### 1.3 Sensitivity analysis for influential clades (Primates data):
The function `clade_phylm` estimate the impact on model estimates of 
phylogenetic linear regression after removing clades from the analysis.

* Additional arguments:   
__clade.col:__ The name of a column in the provided data frame with clades 
specification (a character vector with clade names).  
__n.species:__ Minimum number of species in the clade in order to include this
clade in the leave-one-out deletion analysis. Default is 5.    
__times:__ The number of repetition for the randomization test  


```{r clade_analysis, echo=T, cache=T, warning=FALSE, fig.height=5, fig.width=10}
# Original data set:
knitr::kable(head(primates$data))
# run analysis:
clade <- clade_phylm(log(sexMaturity) ~ log(adultMass), phy = primates$phy[[1]],
                     data = primates$data, clade.col = "family", times = 99, track = F)
# To check summary results and most influential clades:
summary(clade)
# Visual diagnostics for clade removal:
sensi_plot(clade, "Cercopithecidae")
sensi_plot(clade, "Cebidae")
```

<br><br>

## 2.Phylogenetic uncertainty:

The function `tree_phylm` performs Phylogenetic linear regression evaluating 
uncertainty in trees topology.

* Additional arguments:   
__times:__ Number of times to repeat the analysis with n different trees picked
randomly in the multiPhylo file. If NULL, times = 2

### 2.1 Sensitivity analysis for phylogenetic trees:
```{r tree_analysis, echo=T, cache=T, warning=FALSE}
# This analysis needs a multiphylo file:
class(alien$phy)
alien$phy
# run PGLS accounting for phylogenetic uncertain:
tree <- tree_phylm(log(gestaLen) ~ log(adultMass), phy = alien$phy, 
                   data = alien$data, times = 100, track = F)
# To check summary results:
knitr::kable(summary(tree))
# Visual diagnostics
sensi_plot(tree)
```

<br><br>

## 3.Data uncertainty:

The function `intra_phylm` performs Phylogenetic linear regression evaluating
intraspecific variability. 

* Additional arguments:   
__Vy:__ Name of the column containing the standard deviation or the standard 
error of the response variable. When information is not available for one taxon,
the value can be 0 or NA.  
__Vx:__ Name of the column containing the standard deviation or the standard 
error of the predictor variable. When information is not available for one taxon,
the value can be 0 or NA  
__x.transf:__ Transformation for the response variable (e.g. log or sqrt). Please 
use this argument instead of transforming data in the formula directly.  
__y.transf:__ Transformation for the predictor variable (e.g. log or sqrt). Please 
use this argument instead of transforming data in the formula directly.  
__distrib:__ A character string indicating which distribution to use to generate
a random value for the response and/or predictor variables. Default is normal
distribution: "normal" (function rnorm). Uniform distribution: "uniform" (runif)
Warning: we recommend to use normal distribution with Vx or Vy = standard 
deviation of the mean.      
__times:__ Number of times to repeat the analysis with n different trees picked
randomly in the multiPhylo file. If NULL, times = 2   

### 3.1 Sensitivity analysis for intraspecific variation and measurement error:
```{r intra_analysis, echo=T, cache=T, warning=FALSE}
# run PGLS accounting for intraspecific variation:
intra <- intra_phylm(gestaLen ~ adultMass, phy = alien$phy[[1]], track = F, 
                     data = alien$data, Vy = "SD_gesta", Vx = "SD_mass",
                     times = 100, x.transf = log, y.transf = log)
# To check summary results:
knitr::kable(summary(intra))
# Visual diagnostics
sensi_plot(intra)
```

<br><br>

## 4. Aditional functions

### 4.1 Phylogenetic signal for missing data

The function `miss.phylo.d` Calculates D statistic (Fritz & Purvis 2010), 
a measure of phylogenetic signal, for missing data. Missingness is recoded into
a binary variable (1=missing, 0=non missing). 

```{r miss.phylo , echo=T, cache=T, warning=FALSE}
# Load caper:
library(caper)
# Load data
data(alien)
knitr::kable(head(alien.data))
data <- alien.data
phy = alien.phy[[1]]

# Test phylogenetic signal for missing data:
homeNAsig <- miss.phylo.d(data, phy, binvar = homeRange)
print(homeNAsig)
plot(homeNAsig)

massNAsig <- miss.phylo.d(data, phy, binvar = adultMass)
print(massNAsig)
plot(massNAsig)
```

### 4.2 Combine data and phylogeny automatically 

The funcion `match_dataphy` combines phylogeny and data to ensure that tips in phylogeny match data
and that observations with missing values are removed.

> This function uses all variables provided in the 'formula' to match data and phylogeny. To avoid cropping the full dataset, 'match_dataphy' searches for NA values only on variables provided by formula. Missing values on other variables, not included in 'formula', will not be removed from data.

```{r match_dataphy , echo=T, cache=T, warning=FALSE }
# Load data:
data(alien)
# Match data and phy based on model formula:
comp.data <- match_dataphy(gestaLen ~ homeRange, data = alien$data, alien$phy[[1]])
# With a `multiphylo` tree:
comp.data2 <- match_dataphy(homeRange ~ homeRange, data = alien$data, alien$phy)
# Check combined data:
knitr::kable(comp.data$data)
# Check phy:
plot(comp.data$phy)
# See species dropped from phy or data:
comp.data$dropped
```

